---
title: "Explainable AI (XAI)"
author: "Christoph Molnar"
date: "July 05, 2019"
output: 
  ioslides_presentation:
    widescreen: false
    smaller: false
css: ../styles.css
---


```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
  warning = FALSE,
  message = FALSE, 
  fig.align='center', 
  out.width = '90%',
  dev = "CairoPNG")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 15))
```


```{r, child="../chunks/story-wines.Rmd"}
```

# Why interpretability? 

## {.center data-background=../images/explain.png data-background-size=contain}


```{r child = "../chunks/book.Rmd"}
```


# Backup

## Units in Wine dataset

- fixed acidity g(tartaric acid)/dm^3^
- volatile acidity: g(acetric acid/dm^3^)
- citric acid: g/dm^3^
- residual sugar: g/dm^3^
- chlorides: g(sodium chloride)/dm^3^
- free sulfur dioxide: mg/dm^3^
- total sulfur dioxide: mg/dm^3^
- density> g/cm^3^
- pH
- sulphates: g(postassium sulphate) / dm^3^
- alcohol vol.%


```{r, child="../chunks/tools.Rmd"}
```
